Towards an open-source landscape for 3D CSEM modelling

### 2021-04-15 manuscript/Manuscript.tex
Revision II

\documentclass[extra, camera,%
    % onecolumn,   % comment for two columns
    % referee,     % uncomment for double lines
    % final,       % uncomment for changes
]{gji}
\usepackage{timet}

@@

% Own commands
\newcommand{\mr}[1]{\mathrm{#1}}
\newcommand{\emg}[2]{\texttt{emg#1#2}\xspace}
\newcommand{\empymod}{\texttt{empymod}\xspace}
\newcommand{\simpeg}{\texttt{SimPEG}\xspace}
\newcommand{\discretize}{\texttt{discretize}\xspace}
\newcommand{\custem}{\texttt{custEM}\xspace}
\newcommand{\petgem}{\texttt{PETGEM}\xspace}
\newcommand{\ymark}[1]{%
  \begin{tcolorbox}[colback=yellow!40!white]#1\end{tcolorbox}}

% Remove once done
\newcommand{\itodo}[1]{\todo[color=blue!40!white, inline]{\sffamily #1}}
% \newcommand{\mtodo}[1]{\todo[inline]{\sffamily #1}}
\newcommand{\etodo}[1]{\todo[color=blue!40!white]{#1}}
\newcommand{\dul}[1]{\underline{\underline{#1}}}
\newcommand{\ul}[1]{\underline{#1}}
\newcommand{\rmk}[1]{{\color{red}{#1}}}
\newcommand{\ohmm}{\ensuremath{\Omega\,}\text{m}\xspace}

@@

A layered model fails to show the strength of 3D modellers in general and FE codes in particular, and in reality one would not choose a 3D code to compute responses for such a problem. However, it is one of the few examples that can be verified with semi-analytical results, and it is therefore an important first test. The results are shown in Figure~\ref{fig:results-layered}, in the left column the amplitudes and in the right column the phases. The top row shows the actual, semi-analytically computed responses, for which we used \empymod\citep{GEO.17.Werthmuller}. The second and the third row show the relative percentage error of the inline and the broadside receivers, respectively; note that for the 1D case the x-directed electric fields for the two broadside lines for $y=\pm3\,$km are identical.
%
\begin{figure*}
  \centering
  \includegraphics[width=.9\fwidth]{figures/results-layered.png}
  \caption{Results of the layered model: The semi-analytical inline and broadside responses are shown in the top row. The relative percentage error of the four 3D modellers are shown in the second and the third row for the inline and broadside responses, respectively.}
  \label{fig:results-layered}
\end{figure*}
%
The misfit is generally in the order of 1\,\% or less, the biggest misfits are close to the source and where there are rapid phase changes associated to amplitudes having a dip towards zero. The increased errors towards the location of the 200\,m long source are related to the gridding, and finer discretizations (resulting in longer runtime and higher memory consumption) would reduce these errors.

@@

The Marlim oil field is a giant reservoir in a turbidite sandstone horizon in the north-eastern part of the Campos Basin, offshore Brazil, which was discovered in 1985. \cite{BJG.17.Carvalho} created from seismic data and well log data a realistic, three-dimensional resistivity model with vertical transverse isotropy (VTI), called MR3D, which they released under the open creative common license \emph{CC BY 4.0}. \cite{GEO.19.Correa} computed CSEM data for MR3D for six frequencies from 0.125\,Hz to 1.25\,Hz, and released them under the same CC license. To compute the data they used a state-of-the-art code from the industry \citep[][ \emph{SBLwiz} software from \emph{EMGS}]{GEO.07.Maao}. It is therefore, on the one hand, an ideal case to validate our open-source codes against, as it is a complex, realistic model and the data were computed by an industry-proofed code. On the other hand it is impossible to reproduce it exactly, as it is a closed-source code and we cannot know exactly what was done internally. Additionally, that code is a time-domain code obtaining the frequency-domain responses through a transform, and it includes the air layer via a non-local boundary condition at the water-air interface \citep{GEO.10.Mittet}. We compute all the results in this study in the frequency domain, where we model the entire domain including the air layer by putting the boundaries of the computational domain far away.

@@

The responses for all six frequencies and all three electric components are shown in Figure~\ref{fig:results-marlim-responses}, both inline and broadside (we only show the amplitude, as the conclusions for the phase are very similar). The published responses are shown in colour with markers for the frequency, and the responses from our codes are shown beneath in grey colours in this overview plot.
%
\begin{figure*}
  \centering
  \includegraphics[width=.9\fwidth]{figures/results-marlim-responses.png}
  \caption{MR3D comparison between our codes (grey lines) and the published data (coloured lines). The grey lines under the coloured lines are not visible in most areas, meaning that they are very similar. However, there are three notable zones, (1) to (3), which are explained further in the text.}
  \label{fig:results-marlim-responses}
\end{figure*}
%
Two CSEM data-sets were published with MR3D, one containing clean responses, and one responses where realistic noise was added. For the comparison we use the clean data, but we indicate the chosen noise level by the horizontal line at \num{2e-15}\,V/m. It can be seen that the grey lines are not visible for most parts, which means that the data agree well. However, there are a few notable points where this is not the case, and they are marked with numbers in the figure.
\begin{enumerate}
  \item[(1)] There are some noticeable differences for positive offsets of the broadside $E_y$ and $E_z$ components, which are likely related to the bathymetry. The reason why we cannot reproduce the published results exactly is that the code is not open-source, and we do not know in detail what it does internally. See the discussion around Figure~\ref{fig:results-marlim_2published}.
  \item[(2)] The three highest frequencies of the $E_z$ fields become noisy at large offsets for all of our codes (well below any real-world noise level). The reason why the published responses are not noisy is probably because they were computed in the time domain, and transformed to the frequency domain. This explains their smooth behaviour, without saying anything about their accuracy.
  \item[(3)] The responses for the $E_y$ component of the inline acquisition line do not agree at all. In the 1D case, the inline $E_y$ component would be zero, and the only response we can measure here comes from 3D effects. These responses are very low, roughly two orders of magnitude lower than the $E_x$ component. Not the published nor our responses are stable, and any of the responses is as bad as the other. Tiniest differences in meshing or different interpolation algorithms will have a huge effect here.
\end{enumerate}

It is important to note that there is no \emph{true solution}; but the closer that the previously published results and the results obtained from the four codes run here agree, the more confident we can be of the outcomes. This emphasises the importance of comparing 3D results and one of our main objectives, as there is no other way to check the results of 3D codes for complex models. Figure~\ref{fig:results-marlim_2published} shows the normalised differences between the published results and our codes for the three strongest components, the inline $E_x$ and the broadside $E_x$ and $E_y$ components, for three frequencies.
%
\begin{figure*}
  \centering
  \includegraphics[width=.9\fwidth]{figures/results-marlim_2published.png}
  \caption{Normalised differences for all four codes in comparison with the published data, for the inline $E_x$ field and the broadside $E_x$ and $E_y$ fields in the left, middle, and right column, respectively. Shown are the three frequencies 0.125\,Hz, 0.5\,Hz, and 1.0\,Hz in the top, middle, and bottom row, respectively. The results of the two FE codes are almost identical, see Figure~\ref{fig:results-marlim_2ours}, and are therefore combined here.}
  \label{fig:results-marlim_2published}
\end{figure*}
%
A few conclusions can be drawn from this figure: In general the normalised difference is roughly 10\,\% or less. The structured meshes have in some cases a smaller normalised difference, e.g., for negative offsets in the inline and broadside $E_x$ responses, and in other cases the unstructured tetrahedral meshes, e.g., for positive offsets for all broadside $E_y$ responses, and often they have a comparable NRMSD. Some normalised differences have an interesting step-pattern, which is more pronounced for the structured meshes than for the unstructured meshes; this is related to the annotated point (1) in Figure~\ref{fig:model-marlim}. The origin of this zigzag pattern is most likely the bathymetry, as the code of the published data uses the bathymetry information in addition to the resistivity cube. This explains why the codes using the tetrahedra mesh, which use the bathymetry as well, do not show this feature.

@@

%% 1. Results: good agreement %%
Overall, we observed an excellent match between the solutions. The amplitudes and phases of the layered model have mostly a relative error of less than 1--2\,\%. The cross-validation of the results for the block model yields a similar picture, with normalised differences within a few percents. For the large, complex MR3D model it is a bit different. First it can be noted that the misfits are higher for higher frequencies. For the lowest frequency, 0.125\,Hz, the normalised difference is a few percent for inline and broadside $E_x$ components, and up to 10\,\% for the broadside $E_y$ component. For higher frequencies these numbers increase to the order of $\pm10\,$\%. The reason for the higher misfit for MR3D than for the layered or block model lies mostly in the discretization, boundary effects, and interpolation strategy. While one might argue that 5--10\,\% NRMSD is a lot, we argue that this is probably as good as it gets, given that the code with which the responses we compare to were computed is not open-source. This has been shown within the process of this study. Our first attempts were based on the originally published, fine MR3D model. This model is very detailed, but its extent is too small in the x- and y-directions for CSEM computations. In order to represent the fine-scale model on a mesh suitable for computation, each of our codes did its own upscaling and extrapolation. However, we were unsuccessful in our attempt to reproduce the published results.  The reason is that there are many ways how you can extrapolate a model, and without knowing the details of how MR3D represents the model on the computational mesh, we could not re-produce this step. We reached out to the authors of MR3D, and they kindly agreed to also publish the upscaled and extended computational model. It is only with this model that we were able to obtain comparable results.

@@

We compare the computed fields and underlying meshes of four different open-source 3D CSEM codes by means of three increasingly complex resistivity models. A layered marine, anisotropic model is used as first example to verify our results with semi-analytical responses. All codes show excellent accuracy with a relative error in the order of a few percents. Three resistive blocks are added to the layered model to increase complexity in the second example. The cross-validation of the outcomes of the different codes shows a normalised difference of a few percent, hence in a similar range as the relative error in the layered model. The third example is Marlim R3D, a realistic, complex, marine resistivity model. The MR3D model and corresponding CSEM responses are open-source, allowing an independent validation of our results with results from another (closed-source) code. In this scenario the normalised difference is in the order of 5--10\,\%, where the main source of error can be attributed to differences in discretization and handling of the bathymetry. The required runtime and memory consumption is primarily controlled by the used solver. The accuracy, on the other hand, depends to a large degree on the mesh. A proper discretization is therefore key for an accurate result, which makes this step the most time-consuming task, not the actual computation of the responses itself. Validating the correctness of a 3D code is a difficult task, and it is essential to have easily accessible benchmark models with reliable and reproducible solutions. Our study is one example of a collaboration facilitated by open practices including sharing of code and data. We hope that these results may be useful for the entire CSEM community at large, and we invite and encourage the community to make more code, modelling scripts, and results publicly available.

>>>>

\documentclass[a4paper, extra, camera,%
    % onecolumn,   % comment for two columns
    % referee,     % uncomment for double lines
    % final,       % uncomment for changes
]{gji}
\usepackage{timet}

@@

% Own commands
\newcommand{\mr}[1]{\mathrm{#1}}
\newcommand{\emg}[2]{\texttt{emg#1#2}\xspace}
\newcommand{\empymod}{\texttt{empymod}\xspace}
\newcommand{\simpeg}{\texttt{SimPEG}\xspace}
\newcommand{\discretize}{\texttt{discretize}\xspace}
\newcommand{\custem}{\texttt{custEM}\xspace}
\newcommand{\petgem}{\texttt{PETGEM}\xspace}
\newcommand{\ymark}[1]{%
  \begin{tcolorbox}[colback=yellow!40!white]#1\end{tcolorbox}}
\newcommand{\rmk}[1]{{\color{red}{#1}}}
\newcommand{\ohmm}{\ensuremath{\Omega\,}\text{m}\xspace}

@@

A layered model fails to show the strength of 3D modellers in general and FE codes in particular, and in reality one would not choose a 3D code to compute responses for such a problem. However, it is one of the few examples that can be verified with semi-analytical results, and it is therefore an important first test. The results are shown in Figure~\ref{fig:results-layered}, in the left column the amplitudes and in the right column the phases. The top row shows the actual, semi-analytically computed responses, for which we used \empymod\citep{GEO.17.Werthmuller}. The second and the third row show the relative percentage error of the inline and the broadside receivers, respectively; note that for the 1D case the x-directed electric fields for the two broadside lines for $y=\pm3\,$km are identical.
%
\begin{figure*}
  \centering
  \includegraphics[width=.9\fwidth]{figures/results-layered.png}
  \caption{Results of the layered model: The semi-analytical inline and broadside responses are shown in the top row. The relative percentage error of the four 3D modellers are shown in the second and the third row for the inline and broadside responses, respectively.}
  \label{fig:results-layered}
\end{figure*}
%
The misfit is generally in the order of 1\,\% or less, the biggest misfits are on the one hand close to the source and on the other hand where there are rapid phase changes associated to amplitudes having a dip towards zero. The increased errors towards the location of the 200\,m long source are related to the gridding, and finer discretizations (resulting in longer runtime and higher memory consumption) would reduce these errors.

@@

The Marlim oil field is a giant reservoir in a turbidite sandstone horizon in the north-eastern part of the Campos Basin, offshore Brazil, which was discovered in 1985. \cite{BJG.17.Carvalho} created from seismic data and well log data a realistic, three-dimensional resistivity model with vertical transverse isotropy (VTI), called MR3D, which they released under the open creative common license \emph{CC BY 4.0}. \cite{GEO.19.Correa} computed CSEM data for MR3D for six frequencies from 0.125\,Hz to 1.25\,Hz, and released them under the same CC license. To compute the data they used a code from the industry \citep[][ \emph{SBLwiz} software from \emph{EMGS}]{GEO.07.Maao}. It is therefore, on the one hand, an ideal case to validate our open-source codes against, as it is a complex, realistic model and the data were computed by an industry-proofed code. On the other hand it is impossible to reproduce it exactly, as it is a closed-source code and we cannot know exactly what was done internally. Additionally, that code is a time-domain code obtaining the frequency-domain responses through a transform, and it includes the air layer via a non-local boundary condition at the water-air interface \citep{GEO.10.Mittet}. We compute all the results in this study in the frequency domain, where we model the entire domain including the air layer by putting the boundaries of the computational domain far away.

@@

The responses for all six frequencies and all three electric components are shown in Figure~\ref{fig:results-marlim-responses}, both inline and broadside (we only show the amplitude, as the conclusions for the phase are very similar). The published responses are shown in colour with markers for the frequency, and the responses from our codes are shown beneath in grey colours in this overview plot.
%
\begin{figure*}
  \centering
  \includegraphics[width=.9\fwidth]{figures/results-marlim-responses.png}
  \caption{MR3D comparison between our codes (grey lines) and the published data (coloured lines). The grey lines under the coloured lines are not visible in most areas, meaning that they are very similar. However, there are three notable zones, (1) to (3), which are explained further in the text.}
  \label{fig:results-marlim-responses}
\end{figure*}
%
Two CSEM data-sets were published with MR3D, one containing clean responses, and one responses where realistic noise was added. For the comparison we use the clean data, but we indicate the chosen noise level by the horizontal line at \num{2e-15}\,V/m. It can be seen that the grey lines are not visible for most parts, which means that the data agree well. However, there are a few notable points where this is not the case, and they are marked with numbers in the figure.
\begin{enumerate}
  \item[(1)] There are some noticeable differences for positive offsets of the broadside $E_y$ and $E_z$ components, which are likely related to the bathymetry. The reason why we cannot reproduce the published results exactly is that the code is not open-source, and we do not know in detail what it does internally. See the discussion around Figure~\ref{fig:results-marlim_2published}.
  \item[(2)] The three highest frequencies of the $E_z$ fields become noisy at large offsets for all of our codes (well below any real-world noise level).
  \item[(3)] The responses for the $E_y$ component of the inline acquisition line do not agree at all. In the 1D case, the inline $E_y$ component would be zero, and the only response we can measure here comes from 3D effects. These responses are very low, roughly two orders of magnitude lower than the $E_x$ component. Not the published nor our responses are stable, and any of the responses is as bad as the other. Tiniest differences in meshing or different interpolation algorithms will have a huge effect here.
\end{enumerate}

It is important to note that there is no \emph{true solution}; but the closer that the previously published results and the results obtained from the four codes run here agree, the more confident we can be of the outcomes. This emphasises the importance of comparing 3D results and one of our main objectives, as there is no other way to check the results of 3D codes for complex models. Figure~\ref{fig:results-marlim_2published} shows the normalised differences between the published results and our codes for the three strongest components, the inline $E_x$ and the broadside $E_x$ and $E_y$ components, for three frequencies.
%
\begin{figure*}
  \centering
  \includegraphics[width=.9\fwidth]{figures/results-marlim_2published.png}
  \caption{Normalised differences for all four codes in comparison with the published data, for the inline $E_x$ field and the broadside $E_x$ and $E_y$ fields in the left, middle, and right column, respectively. Shown are the three frequencies 0.125\,Hz, 0.5\,Hz, and 1.0\,Hz in the top, middle, and bottom row, respectively. The results of the two FE codes are almost identical, see Figure~\ref{fig:results-marlim_2ours}, and are therefore combined here.}
  \label{fig:results-marlim_2published}
\end{figure*}
%
A few conclusions can be drawn from this figure: In general the normalised difference is roughly 10\,\% or less. The structured meshes have in some cases a smaller normalised difference, e.g., for negative offsets in the inline and broadside $E_x$ responses, and in other cases the unstructured tetrahedral meshes, e.g., for positive offsets for all broadside $E_y$ responses, and often they have a comparable NRMSD. Some normalised differences have an interesting step-pattern, which is more pronounced for the structured meshes than for the unstructured meshes; this is related to the annotated point (1) in Figure~\ref{fig:model-marlim}. The origin of this zigzag pattern is most likely the bathymetry, as the code of the published data uses the bathymetry information in addition to the resistivity cube. This explains why the codes using the tetrahedra mesh, which use the bathymetry as well, do not show this feature. None of our codes has currently special interpolation routines implemented to take particular care of a source injection close to a dipping interface, such as the bathymetry, with discontinuities in the normal electrical field and the normal derivatives of both electric and magnetic tangential fields. This effect alone can potentially contribute several percents of NRMSD, as shown by \mbox{\cite{GJI.15.Shantsev}}.

@@

%% 1. Results: good agreement %%
Overall, we observed an excellent match between the solutions. The amplitudes and phases of the layered model have mostly a relative error of less than 1--2\,\%. The cross-validation of the results for the block model yields a similar picture, with normalised differences within a few percents. For the large, complex MR3D model it is a bit different. First it can be noted that the misfits are higher for higher frequencies. For the lowest frequency, 0.125\,Hz, the normalised difference is a few percent for inline and broadside $E_x$ components, and up to 10\,\% for the broadside $E_y$ component. For higher frequencies these numbers increase to the order of $\pm10\,$\%. The reason for the higher misfit for MR3D than for the layered or block model lies mostly in the discretization, boundary effects, and interpolation strategies for the model, the source distribution on the mesh, and to obtain the response at receiver locations. While one might argue that 5--10\,\% NRMSD is a lot, we argue that this is probably as good as it gets, given that the code with which the responses we compare to were computed is not open-source. This has been shown within the process of this study. Our first attempts were based on the originally published, fine MR3D model. This model is very detailed, but its extent is too small in the x- and y-directions for CSEM computations. In order to represent the fine-scale model on a mesh suitable for computation, each of our codes did its own upscaling and extrapolation. However, we were unsuccessful in our attempt to reproduce the published results.  The reason is that there are many ways how you can extrapolate a model, and without knowing the details of how MR3D represents the model on the computational mesh, we could not re-produce this step. We reached out to the authors of MR3D, and they kindly agreed to also publish the upscaled and extended computational model. It is only with this model that we were able to obtain comparable results.

@@

We compare the computed fields and underlying meshes of four different open-source 3D CSEM codes by means of three increasingly complex resistivity models. A layered marine, anisotropic model is used as first example to verify our results with semi-analytical responses. All codes show excellent accuracy with a relative error in the order of a few percents. Three resistive blocks are added to the layered model to increase complexity in the second example. The cross-validation of the outcomes of the different codes shows a normalised difference of a few percent, hence in a similar range as the relative error in the layered model. The third example is Marlim R3D, a realistic, complex, marine resistivity model. The MR3D model and corresponding CSEM responses are open-source, allowing an independent validation of our results with results from another (closed-source) code. In this scenario the normalised difference is in the order of 5--10\,\%, where the main source of error can be attributed to differences in discretization and handling of the bathymetry with corresponding advanced interpolations of the electromagnetic fields. The required runtime and memory consumption is primarily controlled by the used solver. The accuracy, on the other hand, depends to a large degree on the mesh. A proper discretization is therefore key for an accurate result, which makes this step the most time-consuming task, not the actual computation of the responses itself. Validating the correctness of a 3D code is a difficult task, and it is essential to have easily accessible benchmark models with reliable and reproducible solutions. Our study is one example of a collaboration facilitated by open practices including sharing of code and data. We hope that these results may be useful for the entire CSEM community at large, and we invite and encourage the community to make more code, modelling scripts, and results publicly available.
